---
title: "Mapping the occurence of Cape Fur Seal carcasses"
author: "Muhammad Uzair Davids"
date: "2025-02-27"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## **Load in the packages**

```{r loading packages}
library(rinat)         # iNaturalist data
library(sf)            # Spatial data handling
library(ggplot2)       # Static maps
library(ggspatial)     # Static maps
library(leaflet)       # Interactive maps
library(rnaturalearth) # Base map data
library(terra)         # Raster processing (if needed)
library(tidyverse)     # Data cleaning
library(dplyr)         # Data wrangling
library(readr)         # Reading/writing .csv files
library(stringr)  # Advanced string filtering
```

## **Getting our data from iNat**

Before we can get started, we first need to get our data. We will be using data
from iNaturalist. So to ensure quality data we need to specify that it the data
must be *research grade* and we also only want data from South Africa. As for 
how many results we need to draw from, if you look up [Cape Fur Seal](https://www.inaturalist.org/observations?taxon_id=346947) on iNat, 
4,434 observations show up, so we'll just set `maxresults = ` to 4000 to make 
sure we draw from as many of those observations as possible. 


```{r retrieve iNat data}
# Query iNaturalist for Cape Fur Seal (Arctocephalus pusillus) sightings
data_seal <- get_inat_obs(
  taxon_name = "Arctocephalus pusillus",
  quality = "research", # Only research-grade sightings
  geo = TRUE,           # Must have geolocation
  place_id = 113055,    # Place ID for South Africa in iNaturalist
  maxresults = 4000     # Adjust as needed
)

```

## **Filter observations**

The data currently contained in `data_seal` consists of all research grade
observations of Cape Fur Seals in South Africa, dead or alive, but, we 
specifically want observations of *dead* seals. It's time to *"cull"*. 

The first step is to define the keywords that indicate death in seals. These 
keywords will help us to filter out observations of dead seals. 

```{r define keywords}
# Define keywords indicating death in the seals
death_keywords <- c("dead",
                    "died",
                    "carcass",
                    "deceased",
                    "stranded",
                    "washed up",
                    "lifeless",
                    "corpse")

```

Now that we've set the keywords, we should check the column names to know which
columns to filter for the keywords. 

```{r check column names}
# Check available column names to know which columns to check for 
colnames(data_seal)

```

So, the main columns we would want to check are the `tag_list` and `description`
columns, since these contain any additional information regarding the 
observation. So let's filter these.

```{r filtering the data}
# Convert tags and description to lowercase to avoid case sensitivity
seal_dead <- data_seal %>%
  filter(
    str_detect(tolower(tag_list), 
               paste(death_keywords, collapse = "|")) |  # Check tags
    str_detect(tolower(description), 
               paste(death_keywords, collapse = "|")) # Check description
  ) %>%
  select(id, species_guess, datetime, latitude, longitude, 
         tag_list, description, place_guess)

# Check the first few rows to make sure the data is correct
head(seal_dead)

```


