---
title: "Mapping the occurence of Cape Fur Seal carcasses"
author: "Muhammad Uzair Davids"
date: "2025-02-27"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## **Load in the packages**

```{r loading packages}
library(rinat)             # iNaturalist data
library(sf)                # Spatial data handling
library(ggplot2)           # Static maps
library(ggspatial)         # Static maps
library(leaflet)           # Interactive maps
library(htmltools)         # Interactive maps
library(rnaturalearth)     # Base map data
library(rnaturalearthdata) # Additional features
library(terra)             # Raster processing (if needed)
library(tidyverse)         # Data cleaning
library(dplyr)             # Data wrangling
library(readr)             # Reading/writing .csv files
library(stringr)           # Advanced string filtering
```

## **Getting our data from iNat**

Before we can get started, we first need to get our data. We will be using data
from iNaturalist. So to ensure quality data we need to specify that it the data
must be *research grade* and we also only want data from South Africa. As for 
how many results we need to draw from, if you look up [Cape Fur Seal](https://www.inaturalist.org/observations?taxon_id=346947) on iNat, 
4,434 observations show up, so we'll just set `maxresults = ` to 4000 to make 
sure we draw from as many of those observations as possible (and just hope my
laptop will survive). 


```{r retrieve iNat data}
# Query iNaturalist for Cape Fur Seal (Arctocephalus pusillus) sightings
data_seal <- get_inat_obs(
  taxon_name = "Arctocephalus pusillus",
  quality = "research", # Only research-grade sightings
  geo = TRUE,           # Must have geolocation
  place_id = 113055,    # Place ID for South Africa in iNaturalist
  maxresults = 4000     # Adjust as needed
)

```

## **Filter observations**

The data currently contained in `data_seal` consists of all research grade
observations of Cape Fur Seals in South Africa, dead or alive, but, we 
specifically want observations of *dead* seals. It's time to *"cull"*. 

The first step is to define the keywords that indicate death in seals. These 
keywords will help us to filter out observations of dead seals. 

```{r define keywords}
# Define keywords indicating death in the seals
death_keywords <- c("dead",
                    "died",
                    "carcass",
                    "deceased",
                    "stranded",
                    "washed up",
                    "lifeless",
                    "corpse")

```

Now that we've set the keywords, we should check the column names to know which
columns to filter for the keywords. 

```{r check column names}
# Check available column names to know which columns to check for 
colnames(data_seal)

```

So, the main columns we would want to check are the `tag_list` and `description`
columns, since these contain any additional information regarding the 
observation. So let's filter these.

```{r filtering the data}
# Convert tags and description to lowercase to avoid case sensitivity when filtering
seal_dead <- data_seal %>%
  filter(
    str_detect(tolower(tag_list), 
               paste(death_keywords, collapse = "|")) |  # Check tags
    str_detect(tolower(description), 
               paste(death_keywords, collapse = "|")) # Check description
  ) %>%
  select(id, species_guess, datetime, latitude, longitude, 
         tag_list, description, place_guess)

# Check the first few rows to make sure the data is correct
head(seal_dead)

# Save as CSV in the data file so we have a file with the necessary metadata 
# available for anyone interested
write_csv(seal_dead, "data/seal_dead.csv")

```

## **Convert the data to a spatial object (sf)**

Now that we've got the data, we can save the observations as a spatial object 
that can be plotted. As for the projection of the data, we will set it to WGS84
since the end goal is to map the map interactive, and the `leaflet` package 
requires data to be projected at global projection. 

```{r convert to sf}
# Convert to sf object (WGS84 - EPSG:4326)
seal_dead_sf <- st_as_sf(seal_dead,
                         coords = c("longitude", "latitude"),
                         crs = 4326)

# Save as GeoJSON for GIS applications for anyone who would like to download 
# the file for themselves
st_write(seal_dead_sf, "data/seal_dead.geojson", delete_dsn = TRUE)

```

## **Getting the base map and plotting our points**

So, we have our point data, but we don't have a base map to plot the data on. 
Let's download a base map of South Africa from the `rnaturalearth` package.

```{r download base map}
# Get world map data
world <- ne_countries(scale = "medium", returnclass = "sf")

# Filter for South Africa
south_africa <- world[world$name == "South Africa", ]

# Plot base map with 
ggplot() +
  geom_sf(data = south_africa, fill = "lightgray", color = "black") +
  geom_sf(data = seal_dead_sf, aes(color = "Carcass Sightings"), size = 2) +
  scale_color_manual(values = c("red")) +
  labs(title = "Seal Carcass Sightings in South Africa") +
  theme_minimal()

```

## **Making the map interactive**

To make our map interactive we'll use `leaflet`, this will allow people to zoom
in and out of the map and investigate individual points on the map. 

```{r interactive mapping}
# Create leaflet map
leaflet() %>%
  addTiles() %>%  # Add OpenStreetMap basemap
  addPolygons(data = south_africa, 
              color = "black", 
              fillColor = "lightgray", 
              fillOpacity = 0.5, 
              weight = 1, 
              popup = ~name) %>%
  addCircleMarkers(data = seal_dead_sf, 
                   color = "red", 
                   radius = 3, 
                   popup = ~paste("Location:", place_guess)) %>%
  addLegend(position = "bottomright", colors = "red", labels = "Carcass Sightings") %>%
  addScaleBar(position = "bottomleft") %>%
  addMiniMap(toggleDisplay = TRUE)


```

